---
title: "Extinctions Unit"
author: "Nazerke Sharauova, Latosha Sanchez"
maketitle: true
output: github_document
---

## Introduction

In this module, we are taking a look at if there is another imminent mass extinction at hand. This will be done by estimating the current rate of extinctions across species with the past extincion rate in fossil records. We will be working with non-rectangular data, which are unstructured textual data that must be sorted and shaped into usable datasets for the analysis and mapping of data.  

The loss of biodiversity is a critical problem faced across ecosystems. The inclusion of diverse species across ecosystems heighten their ecosystems, ensuring a natural sustainability for other life. When this balance of diversity falls, a whole cycle may be disrupted, causing critical environmental problems and threatening valuable ecosystem resources, inhibiting the natural process. There are growing concerns for the loss of species and evidence that extinction rates are on the rise.


```{r include=FALSE}
library("tidyverse")
library("httr")
library("jsonlite")
#library("printr")
library("tidyr")
knitr::opts_chunk$set(comment=NA)
```

```{r}
#install.packages("httr")
library(httr)
token = "9bb4facb6d23f48efbf424bb05c0c1ef1cf6f468393bc745d42179ac4aca5fee"
```

The following is to test whether GET works, by requesting the version of the data being retrieved.

```{r}
resp <- httr::GET("https://apiv3.iucnredlist.org/api/v3/version")
#resp <- httr::GET("https://apiv3.iucnredlist.org/api/v3/species/page/0?token=9bb4facb6d23f48efbf424bb05c0c1ef1cf6f468393bc745d42179ac4aca5fee")

version <- httr::content(resp)
version
```
In this chunk of code, we form the list of URLs that will be requested from the IUCN Red List.

```{r}
species_ap <- "https://apiv3.iucnredlist.org/api/v3/species"
page <- "/page/"
page_number <- 0:10
query <- "?token="
token <- "9bb4facb6d23f48efbf424bb05c0c1ef1cf6f468393bc745d42179ac4aca5fee"

all_pages <- paste0 (species_ap, page, page_number, query, token)
all_pages[1:3]
```

We then retrieve all the pages by mapping to the GET request, which were requested from the URL lists above.

```{r}
all_results <- map(all_pages, GET)
```

We then pull the information from the first GET request, then identifying a single record. Another way to retrieve a single record is further given (1) 
```{r}
all_results[[1]]

page0 <- content(all_results[[1]])
page0$result[[1]]

# Optional way to retrieve a record
rec <- all_results [[1]] %>% 
  content() %>% 
  getElement("result") %>% 
  getElement(1)
rec

```
The following code turns the gathered data into a rectangular dataframe which will simplify the use of the information when mapping. Following the transformation, we display the number of rows present and the first five of the dataframe. 
```{r}
id_null <- rec$taxonid[[1]]
id_null[map_lgl(id_null, is.null)] <- NA

row_to_tibble <- function(row) {
  tibble(taxonid = row$taxonid,
         scientific_name = row$scientific_name, 
         category = row$category, 
         taxonomy_class = row$class,
         phylum = row$phylum_name,
         family = row$family_name
         )
  }

get_result <- function(x){
  x %>% content() %>% 
    getElement("result") %>% 
    map_dfr(row_to_tibble)
  }

all_species_df <- all_results %>% map_dfr(get_result) 

nrow(all_species_df)
head(all_species_df)
```

We then chose to narrow down the pool of species to only those that were targeted in the study by Ceballos et al.
```{r}

phylum_dfs <- split(all_species_df, all_species_df$phylum)
chordata_df <- phylum_dfs$CHORDATA

```

The following is used to separate species by their category of extinction. For our report, we want to exclude those that are assumed to be living, and identify the species that are conservatively and/or highly conservatively extinct. We were able to distinguish the categories by their codes which are conveyed as follows:
EX - Highly Conservative
EW - Conservative
PE - Conservative

Any categories that differed from the above were removed. After filtering, we were able to identify the number of species categorized as highly conservative, conservative, and those alive, respectively. 


```{r}
hi_conservative_ex <- chordata_df %>% 
  filter(category == "EX")

conservative_ex <- chordata_df %>% 
  filter(category %in% c("EX","EW","PE")) 

alive <- chordata_df %>%
  filter(!(category %in% c("EX","EW","PE")))

nrow(hi_conservative_ex)
nrow(conservative_ex)
nrow(alive)
```

Below we retrieve the narratives for each of the selected categories. 
```{r}
# Given highly conservative and conservative extinct species, retrieve narratives for each

hi_cons_names <- c(hi_conservative_ex$scientific_name)
cons_names <- c(conservative_ex$scientific_name)

retrieve_narrative <- function(names){
  base <- "https://apiv3.iucnredlist.org"
  narrative <- "/api/v3/species/narrative/"
  urls <- paste0(base, narrative, names, query, token)
  resps <- map(urls, GET)
}

hi_con_resps <- retrieve_narrative(hi_cons_names)
con_resps <- retrieve_narrative(cons_names)

```

The following gives an example of the content in a given row. We will utilized the content of the data to extract corresponding dates to the categories of extinction. 
```{r}
rec <- content(hi_con_resps[[1]])
rec
```

Below we extract the years from the content of the conservative and highly conservative datasets.
```{r}

hi_con_ex_content <- map(hi_con_resps, content)
con_ex_content <- map(con_resps, content)

helper <- function(con){
  result <- con$result[[1]]
  result[map_lgl(result, is.null)] <- NA
  as.data.frame(result)
}

get_max_value <- function(x) {
  if(identical(x, character(0))){
    return(NA)
  }
  max_value <- as.integer(x) %>% max()
  return(max_value)
}

to_narrative_df <- function(resp_content){
  df <- map_dfr(resp_content, helper) %>% 
   select(species_id, rationale) %>%
   mutate(years = stringr::str_extract_all(rationale, "\\d\\d\\d\\d")) %>%
   mutate(extinction_year = map_int(years, get_max_value)) %>%
   drop_na(extinction_year)
  df
}

hi_con_df <- to_narrative_df(hi_con_ex_content)
con_df <- to_narrative_df(con_ex_content)


```

We further analyzed the gathered data by joining the extinction years. From the new dataset, we counted by grouping, evaluating the counts to quantify the cumulative extinctions and join totals by computing percentages. The data is then plotted to visualize the entirety of the data. The background rate has further been included in the comparison of the rates between those from 200 years ago and present.

```{r}

map_chr_safe <- possibly(map_chr, otherwise = NA)

get_period <- function(x) {
  if(is.na(x)) {return(-1)}
  if(x >= 1500 & x <= 1600) {return(1)}
  if(x >= 1600 & x <= 1700) {return(2)}
  if(x >= 1700 & x <= 1800) {return(3)}
  if(x >= 1800 & x <= 1900) {return(4)}
  if(x >= 1900 & x <= 2014) {return(5)}
  return(-1)
}

get_period_range <- function(x) {
  if(is.na(x)) return("Unavailable")
  if(x == 0) return("Before 1500")
  if(x == 1) return("1500-1600")
  if(x == 2) return("1600-1700")
  if(x == 3) return("1700-1800")
  if(x == 4) return("1800-1900")
  if(x == 5) return("1900-2014")
}

analyze_extinctions <- function(narrative_df, chordata_df){
  
  ceballos_classes <- c("MAMMALIA","AVES", "AMPHIBIA", "REPTILIA", "ACTINOPTERYGII", "CEPHALASPIDOMORPHI")

  # Join species data with extinction years from narrative, assign period, and count by grouping
  extinctions_df <- chordata_df %>% 
    left_join(narrative_df, by=c("taxonid" = "species_id")) %>%
    mutate(extinction_period = map_dbl(extinction_year,get_period)) %>%
    group_by(extinction_period, taxonomy_class) %>%
    summarise(num_extinct = n()) %>%
    filter(extinction_period > 0)
  
  # For missing periods per taxonomy_class, introduce zeroes
  for(taxon_class in ceballos_classes){
    for(period in 1:5){
      find_row <- extinctions_df %>%
        filter(taxonomy_class == taxon_class) %>%
        filter(extinction_period == period)
      if(nrow(find_row) == 0){
        extinctions_df[nrow(extinctions_df) + 1, ] = list(period, taxon_class, 0)
      }
    }
  }
  
  # Compute total evaluated counts
  totals_df <- chordata_df %>% 
    left_join(narrative_df, by=c("taxonid" = "species_id")) %>%
    filter(taxonomy_class %in% ceballos_classes) %>%
    group_by(taxonomy_class) %>%
    summarise(total_num = n())

  # Compute cumulative extinctions and join totals
  extinctions_df <- extinctions_df %>%
    arrange(extinction_period) %>%
    group_by(taxonomy_class) %>%
    mutate(cum_num_extinct = cumsum(num_extinct)) %>%
    left_join(totals_df, by=c("taxonomy_class" = "taxonomy_class"))
  
  # Create two new "classes", to match the work by Ceballos et al
  vertebrates_ceballos = c("MAMMALIA", "AVES", "AMPHIBIA")
  other_vertebrates_ceballos = c("REPTILIA", "AMPHIBIA", "ACTINOPTERYGII", "CEPHALASPIDOMORPHI")
  
  for(period in 1:5){
    num_ex <- 0
    cum_num_ex <- 0
    tot <- 0
    for(taxon_class in vertebrates_ceballos){
      tmp <- extinctions_df %>%
        filter(taxonomy_class == taxon_class) %>%
        filter(extinction_period == period)
      num_ex <- num_ex + tmp$num_extinct[[1]]
      cum_num_ex <- cum_num_ex + tmp$cum_num_extinct[[1]]
      tot <- tot + tmp$total_num[[1]]
    }
    extinctions_df[nrow(extinctions_df) + 1, ] = list(period, "VERTEBRATES", num_ex, cum_num_ex, tot)
  }
  
  for(period in 1:5){
    num_ex <- 0
    cum_num_ex <- 0
    tot <- 0
    for(taxon_class in other_vertebrates_ceballos){
      tmp <- extinctions_df %>%
        filter(taxonomy_class == taxon_class) %>%
        filter(extinction_period == period)
      num_ex <- num_ex + tmp$num_extinct[[1]]
      cum_num_ex <- cum_num_ex + tmp$cum_num_extinct[[1]]
      tot <- tot + tmp$total_num[[1]]
    }
    extinctions_df[nrow(extinctions_df) + 1, ] = list(period, "OTHER_VERTEBRATES", num_ex, cum_num_ex, tot)
  }

  # Compute cumulative percentages
  extinctions_df <- extinctions_df %>%
    arrange(extinction_period) %>%
    mutate(cum_percentage = cum_num_extinct/total_num) %>%
    filter(taxonomy_class %in% c("MAMMALIA", "AVES", "VERTEBRATES", "OTHER_VERTEBRATES")) %>%
    mutate(extinction_range = map_chr_safe(extinction_period, get_period_range))

  totals_for_background <- extinctions_df %>%
    filter(extinction_period == 1)
  all_totals <- sum(totals_for_background$total_num)

  # Insert background rate as data
  bg_df <- as_tibble(extinctions_df[nrow(extinctions_df), ])
  cumul <- 0
  for(period in 1:5){
    cumul <- cumul + all_totals/10000*2
    bg_df[nrow(bg_df) + 1, ] = list(period, "BACKGROUND", all_totals/10000*2, cumul, all_totals,cumul/all_totals,get_period_range(period))
  }
  bg_df <- bg_df[-c(1),]
  extinctions_df <- bind_rows(extinctions_df, bg_df)

  # Plot the results
  extinctions_df$cum_percentage = extinctions_df$cum_percentage*100.
  
  extinctions_df %>% 
    ggplot(aes(
       x = extinction_range, 
       y = cum_percentage, 
       color = taxonomy_class, 
       group=as.factor(taxonomy_class))) + 
    geom_line(size = 1) + 
    geom_point(size = 2) +
    scale_color_brewer(palette = "Set1") + 
    theme_bw() + 
    labs(
      title = "Cumulative Vertabrate Species Extinction Rate \n(EX, EW, PE) Over Time", 
         y = "Cumulative Speciece Extintion %", 
         x = "Time Interval", 
         caption = "Conservative estimate of the percentage of the number of species evaluated among mammals.") +
      theme(
    plot.title = element_text(face = 'bold'),
    axis.title = element_text(size = 11, face = 'bold'),
    axis.text = element_text(size = 9),
    plot.caption = element_text(face = 'italic')) +
    scale_color_discrete(name="Taxonomy Class", 
                         breaks=c("AVES",
                                  "BACKGROUND", 
                                  "MAMMILIA", 
                                  "OTHER_VERTEBRATES", 
                                  "VERTEBRATES"),
                         labels=c("Birds", 
                                  "Background", 
                                  "Mammals", 
                                  "Other Vertebrates", 
                                  "Vertebrates"))

}


analyze_extinctions(hi_con_df, chordata_df)

analyze_extinctions(con_df, chordata_df)
```

The graph above, exhibits the extinction rates across vertebrates. It depicts the image that extinction rates have dramatically increased over the past 200 years. To estimate the extinction rates, we gathered available data, and the extinction or possible extinction, on the desired vertebrate species from the 2014 IUCN Red List. 

## Conclusion
Through our analysis, we confirm that the data is comparable to the study conducted by the given research by Ceballos. Thus we can suggest that the extinction rate has continued, and continues to rise at an alarming rate.  

## Citations
Ceballos, G., Ehrlich, P. R., Barnosky, A. D., García, A., Pringle, R. M., & Palmer, T. M. (2015, June 19). Accelerated modern human–induced species losses: Entering the sixth mass extinction. Science Advances, 1, no. 5, e140023. Retrieved from Environmental Sciences (DOI: 10.1126/sciadv.1400253).





